---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
This report was created for the Coursera Reproducible Research Course.
The initial repository was forked from https://github.com/rdpeng/RepData_PeerAssessment1 .
The figures are stored in default RStudio location  (PA1_template_files\figure-html)

The environment locale has to be set to EN.

```{r}
Sys.setlocale("LC_TIME", "English")

```

In the data processing I am using following libraries.

```{r}
library(plyr)
library(lattice)

```

## Loading and preprocessing the data
The dataset [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) for this assignement was downloaded on `r Sys.Date()`.


```{r}
download.file('https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip'
              , "activity_monitoring_data.zip")
unzip("activity_monitoring_data.zip")
allData <- read.csv("activity.csv")

```

I also add a column with a date/time to the dataset.

```{r}
times <- paste(allData$interval %/% 100, allData$interval %% 100, "00", sep = ":")
allData$dateTime <- strptime(paste(allData$date, times) , "%Y-%m-%d %H:%M:%S")

```

## What is mean total number of steps taken per day?
I use the plyr library and make a summary dataset containing the data I might need. For the mean and median calculation I remove the NA values.

```{r}
dayData <- ddply(allData, ~date, summarize, 
                 mean=mean(steps),
                 sd=sd(steps), 
                 median=median(steps, na.rm = TRUE), 
                 sum=sum(steps))

```

Now calculate the Median and Mean
```{r}
totalMean <- mean(dayData$sum, na.rm = TRUE)
totalMedian <- median(dayData$sum, na.rm = TRUE)
```

```{r}
totalMean
```

```{r}
totalMedian
```

And plot the histogram of total number of steps per day

```{r}
hist(dayData$sum, breaks = 50, 
     main = "Histogram of total number of steps per day", 
     xlab = "Steps per day", 
     ylab = "Amount of days", 
     col = "blue", 
     xlim = c(0,25000))
```

The average number of steps per day is **`r as.integer(totalMean)`** with the median of **`r as.integer(totalMedian)`**.

## What is the average daily activity pattern?
Lets calculate and graph the daily pattern. Here is the summary over an interval with its maximum

```{r}
intervalData <- ddply(allData, ~interval, summarize, 
                      mean=mean(steps, na.rm = TRUE), 
                      sd=sd(steps, na.rm = TRUE), 
                      median=median(steps, na.rm = TRUE), 
                      sum=sum(steps, na.rm = TRUE))

maxStepsInInterval <- intervalData[which.max(intervalData$mean),]$interval

maxStepsInInterval
```

The plot for the daily activity follows

```{r}
plot(intervalData$interval, intervalData$mean, type = "l", 
     xlab = "5 min interval", 
     ylab = "Average across all Days", 
     main = "Average number of steps in 5 min Interval", 
     col = "blue")
```

**`r as.integer(maxStepsInInterval)`** is the interval containing the most steps


## Imputing missing values

How many NAs are in the dataset?

```{r}
totalNAs <- sum(is.na(allData$steps))
totalNAs
```

In this step I replace the NA with a mean for the specific interval.

```{r}
allDataWithoutNAs <- allData

for (i in 1:nrow(allDataWithoutNAs))
    if (is.na(allDataWithoutNAs[i,]$steps)) 
        allDataWithoutNAs[i,]$steps <-
    round(intervalData[intervalData$interval ==
                           allDataWithoutNAs[i,]$interval, ]$mean)
```

And prepare summary for the graph + compute the Median and Mean of the "corrected" dataset

```{r}
dayDataWithoutNAs <- ddply(allDataWithoutNAs, ~date, summarize, 
                           mean=mean(steps), 
                           sd=sd(steps), 
                           median=median(steps, na.rm = TRUE), 
                           sum=sum(steps))

totalMeanWithoutNAs <- mean(dayDataWithoutNAs$sum, na.rm = TRUE)
totalMedianWithoutNAs <- median(dayDataWithoutNAs$sum, na.rm = TRUE)
```

```{r}
totalMeanWithoutNAs
```

```{r}
totalMedianWithoutNAs
```

Last but not least the histogram of the total amount of steps per day.

```{r}
hist(dayDataWithoutNAs$sum, breaks = 50, 
     main = "Histogram of total number of steps per day - modified", 
     xlab = "Steps per day", 
     ylab = "Amount of days", 
     col = "red", 
     xlim = c(0,25000))
```

There are in total **`r totalNAs`** missing values (NAs) in the original dataset.
My preference would be, not to fill the missing number, however for this assignement I will use the mean calculated for the specific interval over all days.

The average number of steps per day in the modified dataset is **`r as.integer(totalMeanWithoutNAs)`** with the median of **`r as.integer(totalMedianWithoutNAs)`**.


## Are there differences in activity patterns between weekdays and weekends?

A summary dataset with weekday/end column is created

```{r}
weekend <- ifelse(weekdays(allData$dateTime) %in% 
                      c('Saturday','Sunday'), 'Weekend', 'Weekday')
allData$weekend <- weekend

intervalDataWeek <- ddply(allData, c('interval', 'weekend'), summarize, 
                          mean=mean(steps, na.rm = TRUE))
```

And the final plots with the mean of steps per weekday/end

```{r}
xyplot(mean ~ interval | weekend, intervalDataWeek, type = "l", 
       layout = c(1, 2), 
       xlab = "Interval", 
       ylab = "Average number of steps")
```
